Modeling the impact of IT- and AI-related emissions in a university setting

### 2025-08-05 doc/paper.tex
Related work

Information technology accounts for over 5\% of global electricity consumption and at least 1.7\% of global carbon emissions~\cite{wb:itu:ict}. At the University of Zürich (UZH), which with over 38,000 students and staff is the largest Swiss university, IT in 2024 accounted for about 1.5 M\tcoe\ (megatons of carbon dioxide equivalent) of greenhouse gas (GHG) emissions, approximately 6.3\% of total university emissions in 2024~\cite{uzh:sustainability:report}. The cited report takes into account university-owned workstations, servers, and audiovisual equipment, but does include other IT-related activities such as networking, cloud computing, or the growing use of artificial intelligence, and in particular the adoption of generative AI and large language models (LLMs).

@@

The energy consumption of LLMs is being studied extensively~\cite{budennyy2022eco2ai,castano2023exploring,devries2023growing,epoch2024optimallyallocatingcomputebetweeninferenceandtraining,gowda2024watt,harding2024watts,heguerte2023estimate,luccioni2022estimating,luccioni2023counting,patterson2021carbon,rodriguez2024evaluating,tripp2024measuring,epoch2025howmuchenergydoeschatgptuse}.

@@

Second, management of physical IT assets (workstations, monitors, etc.) is probably an underappreciated and underutilized lever for emissions reduction. Total emissions for these devices are dominated by manufacturing and disposal, so extending their life by a certain amount leads to an almost reciprocal reduction in emissions. For example, if we start the model with default parameters and then increase laptop lifetime from 4 to 6 years and mobile phone lifetime from 3 to 4 years, total campus emissions decrease from 3.9 M\tcoe\ to 3.2 M\tcoe, an 18\% reduction!

@@

@@

>>>>

Information technology accounts for over 5\% of global electricity consumption and at least 1.7\% of global carbon emissions~\cite{wb:itu:ict}. At the University of Zürich (UZH), which with over 38,000 students and staff is the largest Swiss university, IT in 2024 accounted for about 1.5 k\tcoe\ (megatons of carbon dioxide equivalent) of greenhouse gas (GHG) emissions, approximately 6.3\% of total university emissions in 2024~\cite{uzh:sustainability:report}. The cited report takes into account university-owned workstations, servers, and audiovisual equipment, but does include other IT-related activities such as networking, cloud computing, or the growing use of artificial intelligence, and in particular the adoption of generative AI and large language models (LLMs).

@@

@@

Second, management of physical IT assets (workstations, monitors, etc.) is probably an underappreciated and underutilized lever for emissions reduction. Total emissions for these devices are dominated by manufacturing and disposal, so extending their life by a certain amount leads to an almost reciprocal reduction in emissions. For example, if we start the model with default parameters and then increase laptop lifetime from 4 to 6 years and mobile phone lifetime from 3 to 4 years, total campus emissions decrease from 3.9 k\tcoe\ to 3.2 k\tcoe, an 18\% reduction!

@@

Emissions modeling of information technology at organizational scales remains underdeveloped. Universities seem to lack standardized frameworks for measuring IT-related carbon footprints, and our work is in part an effort to help close this gap.

Many carbon footprint studies treat IT as a minor subcomponent, despite IT infrastructure representing 5-10\% of total institutional emissions. For example, \textcite{herth:quantifying:2022} make a comprehensive carbon assessment of TU Delft and find that ICT constitutes 8\% of emissions, but do not break that down further, making it difficult to know how to improve. \textcite{helmers:universities:2021} exclude IT impacts from their 20-university comparison study, claiming that ``almost all universities worldwide will be working with quite similar IT technology''. \textcite{kiehle:2023} include IT equipment in a carbon accounting of the University of Oulu but limit themselves to lifecycle analysis of equipment purchased in the last 3 years.

We are not alone in struggling with data availability problems, including incomplete asset inventories, sparse energy consumption data, etc. \textcite{jurgens:2023} writes that ``data availability is the main limiting factor for a holistic assessment'' of higher education institutions (not limited to IT).

Efforts to standardize IT emissions measurement and reporting are advancing, \eg, through standards such as ITU-T L.1420, but they remain fragmented~\cite{wb:itu:ict}. Our own research on embodied and operational emissions revealed massive variability in carbon reporting among manufacturers and across studies: system boundaries range from cradle-to-gate through cradle-to-grave, functional units vary from per-device to per-computation, and values for the same parameter might vary by a factor of 2 or more.

In this study, our analysis has been bottom up, building a model of the university's IT emissions from individual estimates of devices, queries, etc. There are important tradeoffs between bottom-up precision and top-down comprehensiveness~\cite{hilty:2009:lca}. Bottom-up approaches provide higher accuracy for specific sources and transparency through clear traceability. They also make it easier to identify specific improvement areas. However, they require extensive primary data collection, require up-front decisions about what components to include or not, and may miss system-wide effects. Conversely, top-down approaches offer efficiency and a broader system perspective, but sacrifice the precision and granularity necessary for operational decision-making.

The field of AI is evolving rapidly and the main AI vendors are secretive about the architecture and scale of their models. This has led to many efforts to estimate the footprint of LLMs, from academic papers to blog posts~\cite{devries2023growing,heguerte2023estimate,luccioni2022estimating,ritchie:chatgpt,epoch2025howmuchenergydoeschatgptuse}. More generally, the energy consumption of LLMs is being studied extensively~\cite{budennyy2022eco2ai,castano2023exploring,epoch2024optimallyallocatingcomputebetweeninferenceandtraining,gowda2024watt,harding2024watts,luccioni2023counting,patterson2021carbon,rodriguez2024evaluating,tripp2024measuring}. \textcite{lacoste:mlco2} and \textcite{faiz2024llmcarbon} have developed more systematic approaches to estimate the footprint of models based on factors like electricity emissions intensity, hardware specifications, and network architecture.

In a provocative and maybe dystopian paper, \textcite{tomlinson:2024:carbon} argue that AI has lower emissions than humans for various writing and illustrating tasks. Be that as it may, it would fall into the category of indirect effects of AI, which we do not attempt to cover in this paper.

@@

The proposed research addresses multiple critical gaps: developing consistent methodological frameworks for institutional IT and AI assessment, creating granular modeling approaches that capture technological heterogeneity, integrating embodied and operational emissions across full technology lifecycles, and establishing standardized metrics for institutional comparison and benchmarking. By synthesizing insights from LCA methodologies, AI emissions research, and institutional sustainability practices, this work can contribute substantially to both academic understanding and practical institutional climate action capabilities.
